---
title: "Ceratitis field experiment .Rmd script"
author: "Mariano Ordano"
date: "10 February 2026"
output:
  pdf_document: default
  html_document: default
  word_document: default
editor_options: 
  chunk_output_type: console
---

## Data management

```{r message=T, warning=T, include=T}
knitr::opts_chunk$set(echo = T)
library(readxl)
sc <- read_excel("2026-Cc-cage-experiment.xlsx",sheet=4,col_names=T)
names(sc)
```

### Variable types

```{r message=T, warning=T, include=T}
knitr::opts_chunk$set(echo = T)
str(sc)
sc$trial<-as.factor(sc$trial)
sc$device<-as.factor(sc$device)
sc$treatment<-as.factor(sc$treatment)
sc$eAbbott<-as.numeric(sc$eAbbott)

str(sc)
```

## Data Analysis

## Statistical summary

```{r message=T, warning=T, include=T}
knitr::opts_chunk$set(echo = T)
sc<-as.data.frame(sc)
names(sc)
library(doBy)
sume<-summaryBy(pupae+noeme+malef+femalef+flies+malep+femalep+paras+benef+pbenef+pnoem+pflies+pparas~treatment,FUN=c(mean,sd,median,max),data=sc)
#write.table(sume,"sumsc.txt",quote=F,dec=".",row.names=F) #google
#write.table(sume,"sumsc.txt",quote=F,dec=",",row.names=F)

```

Abbott´s effectiveness calculations

```{r message=T, warning=T, include=T}
knitr::opts_chunk$set(echo = T)
names(sc)
dfa<-subset(sc,device=="D1")
dfa<-subset(dfa,!treatment=="Control")
head(dfa)
str(dfa)
dfa<-as.data.frame(dfa)
sumabbott<-summaryBy(eAbbott~treatment,FUN=c(mean,sd,median,max),data=dfa,na.rm=T)
sumabbott
#write.table(sumabbott,"sumabbott.txt",quote=F,dec=".",row.names=F)

```

### Estimation of treatment effects

```{r message=T, warning=T, include=T}
knitr::opts_chunk$set(echo = T)

# PLOT TIME SERIES GGPLOT2
library(lubridate) # for working with dates
library(ggplot2)  # for creating graphs
library(scales)   # to access breaks/formatting functions
library(gridExtra) # for arranging plots
library(grid)   # for arrangeing plots
library(dplyr)  # for subsetting by season
library(reshape2)
```

#### Benefit

```{r message=T, warning=T, include=T}
knitr::opts_chunk$set(echo = T)
names(sc)
pbeneficio <-aov(pbenef~treatment,sc)

summary(pbeneficio)
shapiro.test(residuals(pbeneficio))

tb <- TukeyHSD(pbeneficio)
tb
```

Table. TukeyHSD. Proportion of benefit.

```{r message=TRUE, warning=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = T)

library(flextable)
# salida de TukeyHSD = tb
tb_df <- as.data.frame(tb$treatment)
tb_df$Comparacion <- rownames(tb_df)
tb_df <- tb_df[, c("Comparacion", "diff", "lwr", "upr", "p adj")]
tb_df[, 2:5] <- round(tb_df[, 2:5], 7)

tabla_word <- flextable(tb_df) %>%
  set_header_labels(
    Comparacion = "Comparación",
    diff = "Diferencia",
    lwr = "Lím. Inferior (95%)",
    upr = "Lím. Superior (95%)",
    "p adj" = "p-valor ajustado"
  ) %>%
  theme_booktabs() %>%
  autofit()

tabla_word

library(officer)
save_as_docx(tabla_word, path = "pbeneficio_tabla_tukey.docx")

```

#### Figure 1. Proportion of benefit (letters follow Tukey HSD differences,  p \< 0.01).

```{r message=F, warning=F, include=T}
knitr::opts_chunk$set(echo = T)
library(gdata)
levels(sc$treatment)
sc$treatment <- factor(sc$treatment, 
                           levels = c("SITBC", "SIT", "BC", "Control"))

library(multcompView)
letras_tukey <- multcompLetters4(pbeneficio, tb)
mis_letras <- letras_tukey$treatment$Letters

bp <- boxplot(pbenef~droplevels(treatment), data=subset(sc, treatment %in%  c('Control','BC','SIT','SITBC')),col=c("gray","green","cyan","violet"), xlab ="Treatments", ylab="Proportion of Benefit", las=1,main="", cex.lab=1.5, cex.axis=1.5, pars=list(boxwex=0.6,staplewex=0.5,outwex=0.5),
               ylim = c(min(sc$pbenef), max(sc$pbenef) * 1.15))

posiciones_y <- bp$stats[4, ] * 1.08
posiciones_y[1] <- bp$stats[4, 1] * 1.12  # grupo 1
posiciones_y[2] <- bp$stats[4, 2] * 2.85  # grupo 2
posiciones_y[3] <- bp$stats[4, 3] * 1.2  # grupo 3 
posiciones_y[4] <- bp$stats[4, 4] * 2.65  # grupo 4
text(x = 1:4,
     y = posiciones_y, #bp$stats[5, ] * 1.1,
     labels = mis_letras, cex = 2, font = 1)

```

#### Abbott's effectiveness

```{r message=T, warning=T, include=T}
knitr::opts_chunk$set(echo = T)
names(sc)
dfa<-subset(sc,device=="D1")
dfa<-subset(dfa,!treatment=="Control")
head(dfa)
str(dfa)
dfa<-as.data.frame(dfa)

```

```{r message=T, warning=T, include=T}
knitr::opts_chunk$set(echo = T)

a1<-aov(eAbbott~treatment,dfa)

summary(a1)
shapiro.test(residuals(a1))

ta <- TukeyHSD(a1)

```

Table. TukeyHSD. Abbott's effectiveness.

```{r message=TRUE, warning=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = T)

library(flextable)
ta_df <- as.data.frame(ta$treatment)
ta_df$Comparacion <- rownames(ta_df)
ta_df <- ta_df[, c("Comparacion", "diff", "lwr", "upr", "p adj")]
ta_df[, 2:5] <- round(ta_df[, 2:5], 7)

atabla_word <- flextable(ta_df) %>%
  set_header_labels(
    Comparacion = "Comparación",
    diff = "Diferencia",
    lwr = "Lím. Inferior (95%)",
    upr = "Lím. Superior (95%)",
    "p adj" = "p-valor ajustado"
  ) %>%
  theme_booktabs() %>%
  autofit()

atabla_word

library(officer)
save_as_docx(atabla_word, path = "eAbbott_tabla_tukey.docx")

```

#### 

#### Figure 2. Abbott´s effectiveness (letters follow Tukey HSD differences,  p \< 0.05).

```{r message=F, warning=F, include=T}
knitr::opts_chunk$set(echo = T)
library(gdata)
levels(dfa$treatment)

library(multcompView)
letras_tukey <- multcompLetters4(a1, ta)
mis_letras <- letras_tukey$treatment$Letters

bpa <- boxplot(eAbbott~droplevels(treatment), data=subset(dfa, treatment %in%  c('BC','SIT','SITBC')),col=c("green","cyan","violet"), xlab ="Treatments", ylab="Effectiveness", las=1,main="", cex.lab=1.5, cex.axis=1.5, pars=list(boxwex=0.6,staplewex=0.5,outwex=0.5),
               ylim = c(min(dfa$eAbbot), max(dfa$eAbbot) * 1.15))

posiciones_y <- bpa$stats[5, ] * 1.1
text(x = 1:4,
     y = posiciones_y, #bp$stats[5, ] * 1.1,
     labels = mis_letras, cex = 2, font = 1)

```

### References

-   R Core Team (2025). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL <https://www.R-project.org/>.)

-   Zuur AF, Ieno EN, Walker NJ, Saveliev AA, Smith G (2009) Mixed effects models and extensions in ecology with R. Dordrecht: Springer.

### Version

20260210 Mariano Ordano maordano\@lillo.org.ar

### AI use

During January 2026, I used the AI Claude to improve the manually written R scripts in 2019, particularly for data subsetting, editing tables and graphs in .Rmd, and exporting files from RStudio in .Rmd for interoperable formats (.docx, .pdf).

### Credits

This script is currently part of the work submitted to the journal Agronomy, entitled ***Combined Effect of Sterile Insect Technique and Augmentative Biological Control Use for Ceratitis capitata Control under Semi-Field Conditions***, authored by Lorena Suárez, Guillermo Sánchez, Mariano Ordano, Fernando Murúa, Segundo Ricardo Núñez-Campero, Flavio Roberto Mello Garcia and Sergio Marcelo Ovruski. Fundings and other recognitions are declared on the paper.
