---
title: "Ceratitis field experiment .Rmd script"
author: "Mariano Ordano"
date: "3 March 2026"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: false
  html_document: default
  word_document: default
editor_options: 
  chunk_output_type: console
---

## Data management

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

```{r chunk1, message=T, warning=T, include=T}
#1
library(readxl)
sc <- read_excel("2026-Cc-cage-experiment.xlsx",sheet=4,col_names=T)
names(sc)
```

### Variable types

```{r chunk2, message=T, warning=T, include=T}
#2
str(sc)
sc$trial<-as.factor(sc$trial)
sc$device<-as.factor(sc$device)
sc$treatment<-as.factor(sc$treatment)
sc$eAbbott<-as.numeric(sc$eAbbott)

str(sc)
```

## Data Analysis

## Statistical summary

```{r chunk3, message=T, warning=T, include=T}
#3
sc<-as.data.frame(sc)
names(sc)
library(doBy)
sume<-summaryBy(pupae+noeme+malef+femalef+flies+malep+femalep+paras+benef+pbenef+pnoem+pflies+pparas~treatment,FUN=c(mean,sd,median,max),data=sc)
#write.table(sume,"sumsc.txt",quote=F,dec=".",row.names=F) #google
#write.table(sume,"sumsc.txt",quote=F,dec=",",row.names=F)

```

Abbott's effectiveness calculations

```{r chunk4, message=T, warning=T, include=T}
#4
names(sc)
dfa<-subset(sc,device=="D1")
dfa<-subset(dfa,!treatment=="Control")
head(dfa)
str(dfa)
dfa<-as.data.frame(dfa)
sumabbott<-summaryBy(eAbbott~treatment,FUN=c(mean,sd,median,max),data=dfa,na.rm=T)
sumabbott
#write.table(sumabbott,"sumabbott.txt",quote=F,dec=".",row.names=F)

```

### Estimation of treatment effects

```{r chunk5, message=T, warning=T, include=T}
#5
library(lme4)        # GLMM
library(lmerTest)    # p-values in linear mixed models
# library(emmeans)     # post-hoc comparisons
library(car)         # Anova tables
library(ggplot2)     # plotting
# library(DHARMa)      # model diagnostics

```

Identifiers for nested factors

```{r chunk6, message=T, warning=T, include=T}
#6
names(sc)
sc$cage_id <- interaction(sc$trial, sc$cage, drop = TRUE)
sc$device_id <- interaction(sc$cage_id, sc$device, drop = TRUE)

```

Checking data structure

```{r chunk7, message=T, warning=T, include=T}
#7
str(sc)
summary(sc)

library(gdata)
levels(sc$treatment)
sc$treatment <- factor(sc$treatment, 
                  levels = c("SITBC", "SIT", "BC", "Control"))
levels(sc$treatment)

# balance
table(sc$treatment, sc$cage)
table(sc$cage, sc$device)
table(sc$cage, sc$trial)

# distribution
hist(sc$pbenef, main = "Proportion of benefit", 
     xlab = "", col = "lightblue")

# Box plots by treatment
ggplot(sc, aes(x = treatment, y = pbenef, fill = treatment)) +
  geom_boxplot() +
  theme_bw() +
  labs(title = "", y = "Proportion of Benefit", x = "Treatment")

# Time trends by treatment
ggplot(sc, aes(x = trial, y = pbenef, color = treatment, group = cage_id)) +
  geom_line(alpha = 0.3) +
  geom_smooth(aes(group = treatment), method = "loess", se = TRUE, linewidth = 1.5) +
  facet_wrap(~treatment) +
  theme_bw() + theme(legend.position = "top") +
  labs(title = "Between Trial Variation by Treatment", y = "Proportion of Benefit", x = "Trial")

```

Mixed-effects model - proportion of benefit

```{r chunk8, message=T, warning=T, include=T}
#8
sc$treatment <- factor(sc$treatment, 
                  levels = c("Control","SITBC", "SIT", "BC"))
levels(sc$treatment)

# 
sc$cage_id <- interaction(sc$trial, sc$treatment, sc$cage, 
                            drop = TRUE, sep = "_")

ms <- lmer(pbenef ~ treatment + 
                     (1|trial) + (1|cage_id),
                     data = sc, REML = TRUE)
summary(ms) # selected model for the paper *

ms2 <- lmer(pbenef ~ treatment + 
                     (1|cage_id),
                     data = sc, REML = TRUE)
summary(ms2)

# model selection
anova(ms, ms2)

# validation
# Gaussian models (lmer)
plot(ms)
qqnorm(resid(ms))
qqline(resid(ms))

# Check random effects
library(lattice)
library(lme4)
ranef(ms)
dotplot(ranef(ms, condVar = TRUE))

# Variance components
VarCorr(ms)

# Type III ANOVA table (for lmer models)
anova(ms, type = 3)


```

Mixed-effects model - Abbott's effectiveness

```{r chunk9, message=T, warning=T, include=T}
#9
names(sc)
dfa$cage_id <- interaction(dfa$trial, dfa$cage, drop = T)

mga <- lmer(eAbbott ~ treatment + (1|cage_id) , data = dfa, REML = TRUE)
summary(mga)

mga2 <- lmer(eAbbott ~ treatment + (1|trial) , data = dfa, REML = TRUE)
summary(mga2) # selected model for the paper *

anova (mga,mga2)

# Type III ANOVA table (for lmer models)
anova(mga2, type = 3)

```

Tables, plots

```{r chunk10, message=T, warning=T, include=T}
#10
library(lubridate) # working with dates
library(ggplot2)  
library(scales)   # breaks/formatting functions
library(gridExtra) # arranging plots
library(grid)   # arrangeing plots
library(dplyr)  # subsetting by season
library(reshape2) # data management

library(sjmisc) # tables
library(sjlabelled)
library(sjPlot)

# devtools::install_github("strengejacke/strengejacke")
set_theme(base = theme_bw(),
          panel.gridcol = "white",
          axis.textcolor.x = "black",
          axis.textcolor.y = "black",
          axis.textcolor = "black",
          axis.title.color = "black",
          axis.title.size = 1,
          axis.textsize = 1)


# tab_model renders HTML tables; only shown in HTML/Word output
if (!knitr::is_latex_output()) {
  tab_model(ms,  transform=NULL, show.se=T, show.ci=NULL, show.stat=T) # selected
  tab_model(ms2, transform=NULL, show.se=T, show.ci=NULL, show.stat=T)
  tab_model(mga, transform=NULL, show.se=T, show.ci=NULL, show.stat=T)
  tab_model(mga2,transform=NULL, show.se=T, show.ci=NULL, show.stat=T) # selected
} else {
  # For PDF: plain text summaries
  print(summary(ms))
  print(summary(ms2))
  print(summary(mga))
  print(summary(mga2))
}

```

#### Tukey HSD - Proportion of Benefit

```{r chunk11, message=T, warning=T, include=T}
#11
names(sc)
pbeneficio <-aov(pbenef~treatment,sc)
tb <- TukeyHSD(pbeneficio)
tb
```

Table. TukeyHSD. Proportion of benefit.

```{r chunk12, message=TRUE, warning=TRUE, include=TRUE}
#12
library(flextable)
# Enable flextable rendering for PDF via knitr
set_flextable_defaults(fonts_ignore = TRUE)

# salida de TukeyHSD = tb
tb_df <- as.data.frame(tb$treatment)
tb_df$Comparacion <- rownames(tb_df)
tb_df <- tb_df[, c("Comparacion", "diff", "lwr", "upr", "p adj")]
tb_df[, 2:5] <- round(tb_df[, 2:5], 7)

tabla_word <- flextable(tb_df) %>%
  set_header_labels(
    Comparacion = "Comparación",
    diff = "Diferencia",
    lwr = "Lím. Inferior (95%)",
    upr = "Lím. Superior (95%)",
    "p adj" = "p-valor ajustado"
  ) %>%
  theme_booktabs() %>%
  autofit()

tabla_word
#library(officer)
#save_as_docx(tabla_word, path = "pbeneficio_tabla_tukey.docx")

```

#### Figure 1. Proportion of benefit (letters follow Tukey HSD differences, p \< 0.01).

```{r chunk13, message=F, warning=F, include=T}
#13
library(gdata)
levels(sc$treatment)
sc$treatment <- factor(sc$treatment, 
                           levels = c("SITBC", "SIT", "BC", "Control"))

library(multcompView)
letras_tukey <- multcompLetters4(pbeneficio, tb)
mis_letras <- letras_tukey$treatment$Letters

bp <- boxplot(pbenef~droplevels(treatment), data=subset(sc, treatment %in%  c('Control','BC','SIT','SITBC')),col=c("gray","green","cyan","violet"), xlab ="Treatments", ylab="Proportion of Benefit", las=1,main="", cex.lab=1.5, cex.axis=1.5, pars=list(boxwex=0.6,staplewex=0.5,outwex=0.5),
               ylim = c(min(sc$pbenef), max(sc$pbenef) * 1.15))

posiciones_y <- bp$stats[4, ] * 1.08
posiciones_y[1] <- bp$stats[4, 1] * 1.12  # grupo 1
posiciones_y[2] <- bp$stats[4, 2] * 2.85  # grupo 2
posiciones_y[3] <- bp$stats[4, 3] * 1.2  # grupo 3 
posiciones_y[4] <- bp$stats[4, 4] * 2.65  # grupo 4
text(x = 1:4,
     y = posiciones_y, #bp$stats[5, ] * 1.1,
     labels = mis_letras, cex = 2, font = 1)

```

#### Tukey HSD - Abbott's effectiveness

```{r chunk14, message=T, warning=T, include=T}
#14
names(sc)
dfa<-subset(sc,device=="D1")
dfa<-subset(dfa,!treatment=="Control")
head(dfa)
str(dfa)
dfa<-as.data.frame(dfa)

```

```{r chunk15, message=T, warning=T, include=T}
#15
a1<-aov(eAbbott~treatment,dfa)
ta <- TukeyHSD(a1)

```

Table. TukeyHSD. Abbott's effectiveness.

```{r chunk16, message=TRUE, warning=TRUE, include=TRUE}
#16
library(flextable)
ta_df <- as.data.frame(ta$treatment)
ta_df$Comparacion <- rownames(ta_df)
ta_df <- ta_df[, c("Comparacion", "diff", "lwr", "upr", "p adj")]
ta_df[, 2:5] <- round(ta_df[, 2:5], 7)

atabla_word <- flextable(ta_df) %>%
  set_header_labels(
    Comparacion = "Comparación",
    diff = "Diferencia",
    lwr = "Lím. Inferior (95%)",
    upr = "Lím. Superior (95%)",
    "p adj" = "p-valor ajustado"
  ) %>%
  theme_booktabs() %>%
  autofit()

atabla_word

#library(officer)
#save_as_docx(atabla_word, path = "eAbbott_tabla_tukey.docx")

```

#### 

#### Figure 2. Abbott's effectiveness (letters follow Tukey HSD differences, p \< 0.05).

```{r chunk17, message=F, warning=F, include=T}
#17
library(gdata)
levels(dfa$treatment)

library(multcompView)
letras_tukey <- multcompLetters4(a1, ta)
mis_letras <- letras_tukey$treatment$Letters

bpa <- boxplot(eAbbott~droplevels(treatment), data=subset(dfa, treatment %in%  c('BC','SIT','SITBC')),col=c("green","cyan","violet"), xlab ="Treatments", ylab="Effectiveness", las=1,main="", cex.lab=1.5, cex.axis=1.5, pars=list(boxwex=0.6,staplewex=0.5,outwex=0.5),
               ylim = c(min(dfa$eAbbott), max(dfa$eAbbott) * 1.15))

posiciones_y <- bpa$stats[5, ] * 1.1
text(x = 1:4,
     y = posiciones_y, #bp$stats[5, ] * 1.1,
     labels = mis_letras, cex = 2, font = 1)

```

### References

-   R Core Team (2025). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL <https://www.R-project.org/>.)

-   Zuur AF, Ieno EN, Walker NJ, Saveliev AA, Smith G (2009) Mixed effects models and extensions in ecology with R. Dordrecht: Springer.

### Version

20260305 Mariano Ordano maordano\@lillo.org.ar

### AI use

From January to March 2026, I used the AI Claude to improve the manually written R scripts in 2019, particularly for data subsetting, editing models, tables and graphs in .Rmd, and exporting files from RStudio in .Rmd for interoperable formats (.docx, .pdf).

### Credits

This script is currently part of the work submitted to the journal Agronomy, entitled ***Combined Effect of Sterile Insect Technique and Augmentative Biological Control Use for Ceratitis capitata Control under Semi-Field Conditions***, authored by Lorena Suárez, Guillermo Sánchez, Mariano Ordano, Fernando Murúa, Segundo Ricardo Núñez-Campero, Flavio Roberto Mello Garcia and Sergio Marcelo Ovruski. Fundings and other recognitions are declared on the paper.
